### 2019-09-27 aistats_paper/sections/experiments.tex
More updates.

We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
Initial experiements were conducted using synthetic data sampled from a known mixture of Gaussians.

@@

To evaluate each method with a real application, we used a random sample of rows from the Gaia DR2 source table~\citep{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
In total there were 2 million rows.
Where data were missing, we set the field to zero and the noise variance to a large value.
We set the projection $R_i$ to the identity matrix for every sample.
This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.

The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
We used a validation set comprising $10\%$ of the rows when developing our experiments.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.

@@

Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
The values are similar, but not exactly comparable, as the effect of regularisation differs for each method.
Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$, and training time as function of mixture components $K$.
Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.

@@

>>>>

We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
Initial experiments were conducted using synthetic data sampled from a known mixture of Gaussians.
We then evaluated the different methods on a real application using a dataset constructed from the Gaia catalogue.

The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes and dataset sizes illustrate how the methods scale.
Full details of our experimental setup for reproducibility are available in the supplementary material.

@@

To evaluate each method with a real application, we used random samples of rows from the Gaia DR2 source table~\citep{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
We set the projection $R_i$ to the identity matrix for every sample.
Where data were missing, we set the field to zero and the noise variance to a large value.
This study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.

@@

In our initial experiments we evaluated the effect of varying the number of mixture components $K$.
A fixed subsample of $2,000,000$ rows was used.
We used a validation set comprising $10\%$ of the rows when developing our experiments.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
The values are similar, but not exactly comparable, as the effect of regularisation differs for each method.
Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$, and training time as function of mixture components $K$.
Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.

@@

\todo{Description of setup.}

\todo{Table of results for different values of N}

\todo{Plot of validation LL against cost-scaled epochs}

\todo{Plot of training time against N}

### 2019-09-27 aistats_paper/sections/methods.tex
More updates.

\subsection{Minibatch Expectation-Maximisation}
\label{sec:minibatch-em}
The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
\begin{align}
r_{ij} &= \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \\ 
\bb_{ij} &= m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \\
B_{ij} &= V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
\end{align}
The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
The expected sufficient statistics are then summed together over the minibatch,
\begin{align}
q_{jt} &= \sum_i r_{ijt}, \\
\bs_{jt} &= \sum_i r_{ijt} \bb_{ijt}, \\
S_{jt} &= \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
\end{align}
Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
\begin{align}
\hat{\phi}_{jt} &= (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\\
\phi_{jt} &= \{q_{jt}, \bs_{jt}, S_{jt} \},\\
\hat{\phi}_{jt} &= \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
\end{align}
Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
\begin{align}
\alpha_{j} &= \frac{\hat{q}_{jt}}{M}, \\
\bm_{j} &= \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \\
V_{j} &= \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
\label{eqn:mstep}
\end{align}
This procedure is repeated with new randomly-ordered minibatches until convergence.
If we set $\lambda=1$ and replace each minibatch with the entire dataset,
% and factorise the update for $V_j$ -- % unclear to someone new?
then the update corresponds to the original batch fitting method.
Numerically however, the update for $V_j$, as written in \eqref{eqn:mstep}, is inadvisable compared to the batch update given in \cite{bovyExtremeDeconvolutionInferring2011}. There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.

@@

In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
We also compute minibatch-specific parameters using exact sums over the minibatch:
\begin{align}
q_b &= \sum_i^M r_i, \\
\bm_b &= \frac{\sum_i^M r_i \bx_i}{q_b},\\
V_b &= \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
\end{align}
We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
\begin{align}
\begin{split}
V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) \\
&+ \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}
\end{split} \\
&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
\end{align}
Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.

>>>>

\subsection{Minibatch Expectation-Maximisation}
\label{sec:minibatch-em}
The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
\begin{align}
r_{ij} &= \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \\ 
\bb_{ij} &= m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \\
B_{ij} &= V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
\end{align}
\todo{Expand out where $\bb_{ij}$ and $B_{ij}$ come from, possibly in background?.}
The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
The expected sufficient statistics are then summed together over the minibatch,
\begin{align}
q_{jt} &= \sum_i r_{ijt}, \\
\bs_{jt} &= \sum_i r_{ijt} \bb_{ijt}, \\
S_{jt} &= \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
\end{align}
Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
\begin{align}
\hat{\phi}_{jt} &= (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\\
\phi_{jt} &= \{q_{jt}, \bs_{jt}, S_{jt} \},\\
\hat{\phi}_{jt} &= \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
\end{align}
Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
\begin{align}
\alpha_{j} &= \frac{\hat{q}_{jt}}{M}, \\
\bm_{j} &= \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \\
V_{j} &= \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
\label{eqn:mstep}
\end{align}
This procedure is repeated with new randomly-ordered minibatches until convergence.
If we set $\lambda=1$ and replace each minibatch with the entire dataset,
% and factorise the update for $V_j$ -- % unclear to someone new?
then the update corresponds to the original batch fitting method.
Numerically however, the update for $V_j$, as written in \eqref{eqn:mstep}, is inadvisable compared to the batch update given in \cite{bovyExtremeDeconvolutionInferring2011}. There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.

@@

In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
We also compute minibatch-specific parameters using exact sums over the minibatch:
\begin{align}
q_b &= \sum_i^M r_i, \\
\bm_b &= \frac{\sum_i^M r_i \bx_i}{q_b},\\
V_b &= \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
\end{align}
We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
\begin{align}
\begin{split}
V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) \\
&+ \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}
\end{split} \\
\begin{split}
&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] \\
&+ \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right]
\end{split} \\
&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
\end{align}
\todo{Fix typesetting here.}
Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
